Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-28 response_to_referees.tex

% 	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future. However, observing apparently-neutral statistics in the case of A/H3N2 is still surprising to us. The case of simulated populations shows that clear signs of selection can be observed by looking at basic features of trajectories (\emph{e.g.} figure S20 of the supplementary material). This is consistent with the theory of evolving population: if the only available information about a mutation is that it has risen in frequency starting from $0$, it has a higher chance on being beneficial than deleterious, and its fixation probability should be higher than what it would be in a neutral scenario. Hence, even with our very simple approach, we observe that predictability of A/H3N2 qualitatively differs from what models of adapting populations would suggest. \\}

\refa{2. The authors examined the predictive power of one predictive
method, the LBI. However, they did not do a systematic comparison of
the different methods that they cite (Morris et al, 2018), which differ in prediction
targets and methods. Therefore, the general conclusions about
predictability, eg. on page 8, in line 47, and on page 9, line 48 are
too sweeping and should be made precisely for those methods looked at
in detail.}

\response{The reviewer is right to point out that our original manuscript did not perform a systematic comparison of different prediction methods.

In a recent model-driven effort (Huddleston et al), we investigate the power of sequence based, antigenically informed, experimentally informed, or phylogenetic (LBI) scores to predict future influenza virus populations.
The LBI was among the highest performing models and we hence used this score in the original manuscript.
But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material (``Ability of fitness models to predict fixation'').
These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
We believe this addition has made the manuscript stronger and we thank both referees for the suggestion.

We also agree with the reviewer that these different methods differ in prediction targets: our results indicate that most of them are not predictive of frequency dynamics and fixation, but they remain predictive of the composition of the future population.
In consequence, we have toned down our conclusions about predictability in the discussion.}
}

### 2020-12-29 response_to_referees.tex

% 	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future. However, observing apparently-neutral statistics in the case of A/H3N2 is still surprising to us. The case of simulated populations shows that clear signs of selection can be observed by looking at basic features of trajectories (\emph{e.g.} figure S20 of the supplementary material). This is consistent with the theory of evolving population: if the only available information about a mutation is that it has risen in frequency starting from $0$, it has a higher chance on being beneficial than deleterious, and its fixation probability should be higher than what it would be in a neutral scenario. Hence, even with our very simple approach, we observe that predictability of A/H3N2 qualitatively differs from what models of adapting populations would suggest. \\}

\refa{2. The authors examined the predictive power of one predictive
method, the LBI. However, they did not do a systematic comparison of
the different methods that they cite (Morris et al, 2018), which differ in prediction
targets and methods. Therefore, the general conclusions about
predictability, eg. on page 8, in line 47, and on page 9, line 48 are
too sweeping and should be made precisely for those methods looked at
in detail.}

\response{The reviewer is right to point out that our original manuscript did not perform a systematic comparison of different prediction methods.

In a recent model-driven effort (Huddleston et al), we investigate the power of sequence based, antigenically informed, experimentally informed, or phylogenetic (LBI) scores to predict future influenza virus populations.
The LBI was among the highest performing models and we hence used this score in the original manuscript.
But we took the suggestion of the reviewer (and reviewer 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material (``Ability of fitness models to predict fixation'').
These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
We believe this addition has made the manuscript stronger and we thank both reviewers for the suggestion.

We also agree with the reviewer that these different methods differ in prediction targets: our results indicate that most of them are not predictive of frequency dynamics and fixation, but they remain predictive of the composition of the future population.
In consequence, we have toned down our conclusions about predictability in the discussion.}

\refa{3. Going through the previous and cited literature, I think the authors should cite some of these works in a more careful way. Specifically, the very related work by Illingworth and Mustonen is not mentioned. The distribution $P_{\delta t}(f|f_0)$ has been at the core of
the method of Strelkowa and Lassig, 2012 (which is cited, but at other
parts of the text), and the same paper also uses a similar simulation model, which should be acknowledged at the appropriate points of the text.}
\response{We now refer to the work by Illingworth and Mustonen in different parts of the text. We also added citations to Strelkowa and L\"assig when discussing the future distribution of frequencies and the simulations.}

@@

\refb{2. Fig 2A and 2B look quite different to me. In the text it appears to me as if these are very similar to the authors. What are the characteristics of the AAs that fix in H1N1? }

\response{Whether 2A or 2B are perceived as similar or not probably depends on how big a deviation from the diagonal one expected a priori.
We now highlight the differences in the text and discuss potential reasons for the different behaviour.  }
\responsedraft{We don't discuss the potential reasons yet!}

\refb{3. Question: what do these results mean for vaccines? How to decide which strains to use for vaccines? This may be known to those who work on Influenza, but for a relative outsider it is not clear. }

\response{
	% Computational prediction of which circulating strains will dominate future influenza populations have been attempted for about 2 decades and have recently become a regular input to the biannual consultations on the influenza vaccine composition.
	% Since performance of predictive models can only be assessed on limited historical data, we believe that our model-free approach clarifies which aspect of influenza evolution are predictable and which ones are not.
	Methods that are useful for designing vaccines are those that predict which circulating strains will dominate future influenza populations.
	Such predictions have been attempted for about 2 decades and have recently become a regular input to the biannual consultations on the influenza vaccine composition.

	On the other hand, our model-free approach is more concerned in understanding which aspects of influenza evolution are predictable and which ones are not.
	The main implications of our results are about the evolutionary dynamics of influenza, and only indirectly about vaccines.  
}

### 2021-01-05 article.tex
discuss h1n1 h3n2 differences

A recent study proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
Furthermore, selection might cause the rapid rise of a novel variant to macroscopic frequencies (observable in a global sample) but its benefit rapidly ``expires'' because competing variants catch up and/or it mediates immune escape only to a small fraction of the population.
These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.

>>>>

A recent study proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
Furthermore, selection might cause the rapid rise of a novel variant to macroscopic frequencies (observable in a global sample) but its benefit rapidly ``expires'' because competing variants catch up and/or it mediates immune escape only to a small fraction of the population \citep{lee_mapping_2019}.
Immunity to A/H1N1pdm might be less variable among humans than to A/H3N2 since A/H1N1pdm has been circulating for only 11 years.
Different age-groups therefore got exposed to more similar A/H1N1pdm variants than their initial exposure to A/H3N2.
These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.
